---
title: "Data Lab: Multiple Mediation Analysis"
format:
  html:
    code-fold: false
---

::: {.callout-info title="Lab Overview"}
In this practical lab, we will perform a **Multiple Mediation Analysis** using Structural Equation Modeling (SEM) with the `lavaan` package. We will simulate a scenario with a binary predictor (X), multiple binary mediators (M), and a binary outcome (Y).
:::

## 1. Data Loading, Dichotomizing & Simulation

We'll start with the UCLA dataset. We will dichotomize our predictor (GRE) and simulate three new binary mediators (passing an essay, interview, and recommendation) that depend on the GRE score.

```{r setup, message=FALSE, warning=FALSE}
# Load required libraries
library(lavaan)
library(dplyr)
library(tidyr) 

# Load the dataset
dt <- read.csv("https://stats.idre.ucla.edu/stat/data/binary.csv")

# A. Dummy coding for the covariates (Rank)
dt <- dt %>%
  mutate(
    rank2 = ifelse(rank == 2, 1, 0),
    rank3 = ifelse(rank == 3, 1, 0),
    rank4 = ifelse(rank == 4, 1, 0)
  )

# B. Make GRE Dichotomous (1 = High GRE [>= 600], 0 = Low GRE [< 600])
dt <- dt %>%
  mutate(
    gre_high = ifelse(gre >= 600, 1, 0)
  )
```

### Simulating Mediators

We simulate the 3 new dichotomous mediators based on the binary `gre_high`. We adjust the logit math. For example, in `prob_essay`: 
- If `gre_high = 0` (Low GRE), prob is $plogis(-1) \approx 27\%$ pass rate.
- If `gre_high = 1` (High GRE), prob is $plogis(-1 + 2) = plogis(1) \approx 73\%$ pass rate.

```{r simulate-mediators}
set.seed(123) 
dt <- dt %>%
  mutate(
    prob_essay = plogis(-1 + 2.0 * gre_high),
    essay_pass = rbinom(n(), size = 1, prob = prob_essay),
    
    prob_interview = plogis(-1.5 + 2.5 * gre_high),
    interview_pass = rbinom(n(), size = 1, prob = prob_interview),
    
    prob_rec = plogis(-0.5 + 1.5 * gre_high),
    rec_pass = rbinom(n(), size = 1, prob = prob_rec)
  )
```

### Preparing Variables for `lavaan`

We define endogenous categorical variables as `ordered` for `lavaan`. 
Note: We do NOT need to make `gre_high` ordered because it is **EXOGENOUS** (no arrows point to it). `lavaan` handles exogenous binary variables perfectly as 0/1 numeric.

```{r prep-lavaan}
dt$admit <- ordered(dt$admit)
dt$essay_pass <- ordered(dt$essay_pass)
dt$interview_pass <- ordered(dt$interview_pass)
dt$rec_pass <- ordered(dt$rec_pass)
```

## 2. Defining the SEM Model

We define the structural paths:
1. $X \rightarrow M$ pathways
2. $M \rightarrow Y$ and $X \rightarrow Y$ pathways
3. We calculate specific indirect effects and total effects using the `:=` operator.

```{r define-model}
model_3_mediators_binaryX <- '
  # The X -> M pathways (Now using gre_high)
  essay_pass ~ a1 * gre_high
  interview_pass ~ a2 * gre_high
  rec_pass ~ a3 * gre_high

  # The M -> Y, X -> Y, and Covariate pathways pointing to Admit
  admit ~ cp * gre_high + b1 * essay_pass + b2 * interview_pass + b3 * rec_pass + gpa + rank2 + rank3 + rank4

  # Specific Indirect Effects
  ind_essay := a1 * b1
  ind_interview := a2 * b2
  ind_rec := a3 * b3

  # Total Effects
  total_indirect := ind_essay + ind_interview + ind_rec
  total_effect := cp + total_indirect
'
```

## 3. Model Fitting & Results

We fit the model using `sem()`. Because we declared our endogenous variables as `ordered`, `lavaan` automatically uses the DWLS (Diagonally Weighted Least Squares) estimator.

```{r model-fit}
# Fit the model 
fit <- sem(model = model_3_mediators_binaryX, data = dt)

# Print full summary
summary(fit, standardized = TRUE, fit.measures = TRUE, rsquare = TRUE)
```

### 3.1 Extracting Indirect Effects

Let's isolate just the indirect effects to compare the mediating pathways.

```{r indirect-effects}
# Extract just the indirect effects to compare
parameterEstimates(fit) %>% 
  filter(op == ":=") %>% 
  select(lhs, est, se, z, pvalue) %>%
  mutate(across(where(is.numeric), ~ round(., 4)))
```

> **Interpretation:** The table above shows the specific indirect effects through the essay, interview, and recommendation. We can determine which pathway carries the most significant mediating effect by observing the `est` (estimate) and `pvalue`.

::: {.callout-tip title="Cross-References"}
* Return to [Topics overview](../../README.md)
:::
