---
title: "Data Lab: Linear Regression Analysis"
format:
  html:
    code-fold: false
---

::: {.callout-info title="Lab Overview"}
In this practical lab, we will perform a comprehensive Linear Regression Analysis. We'll start with Exploratory Data Analysis, fit simple and multiple regression models, verify the OLS assumptions, and visualize our results.
:::

## 1. Setting up the environment

We begin by loading the necessary libraries and our dataset. We'll use the built-in `mtcars` dataset for the simple regression, and generate a synthetic dataset later for multiple regression.

```{r setup, message=FALSE, warning=FALSE}
# Load necessary libraries for data manipulation, visualization, and diagnostic testing
library(datasets)
library(ggplot2) 
library(lmtest) # Required for assumption verification tests (bptest, dwtest, resettest)
library(car)    # Required for multicollinearity test (vif)

# Load the built-in mtcars dataset and preview the first few rows
data(mtcars) 
head(mtcars)
```

## 2. Exploratory Data Analysis (EDA)

Before fitting models, we perform an exploratory analysis to understand the relationships between our variables of interest.

```{r eda, warning=FALSE, message=FALSE}
# Generate summary statistics for the dataset
summary(mtcars) 

# Create a scatter plot of Weight vs. MPG to visually inspect their relationship
ggplot(mtcars, aes(x = wt, y = mpg)) + 
  geom_point(color = "steelblue", size = 3) + 
  labs(title = "Scatter plot of Weight vs. MPG", 
       x = "Weight (1000 lbs)", 
       y = "Miles Per Gallon (MPG)") +
  theme_minimal()

# Calculate the Pearson correlation coefficient between weight and MPG
cor(mtcars$wt, mtcars$mpg)
```

## 3. Simple Linear Regression

We fit a linear regression model to predict `mpg` (Y) based on car weight `wt` (X):
$$Y = \beta_0 + \beta_1 X + \epsilon$$

```{r simple-model}
# Fit the simple linear regression model 
modelo <- lm(mpg ~ wt, data = mtcars) 

# Display the summary of the model
summary(modelo)
```

> **Interpretation:** In the summary output, the intercept is $\beta_0$ and the coefficient under the `wt` column is $\beta_1$ (the slope). The slope is estimated to be -5.34, meaning that for every additional 1000 lbs of weight, the car loses about 5.34 miles per gallon.

### 3.1 Visual Diagnostic Plots

```{r simple-diagnostics, fig.height=8, fig.width=8}
# Generate standard visual diagnostic plots 
# 1. Residuals vs Fitted: Checks for non-linear patterns.
# 2. Normal Q-Q: Checks if residuals are normally distributed.
# 3. Scale-Location: Checks the assumption of constant variance (homoscedasticity).
# 4. Residuals vs Leverage: Identifies influential observations/outliers.
par(mfrow = c(2, 2))
plot(modelo)
```

### 3.2 Formal Verification of OLS Assumptions

Beyond visual diagnostics, we use formal statistical tests to verify the core assumptions.

```{r simple-assumptions}
# 1. Normality of Residuals (Shapiro-Wilk)
shapiro.test(residuals(modelo))

# 2. Homoscedasticity (Breusch-Pagan)
bptest(modelo)

# 3. Independence of Residuals (Durbin-Watson)
dwtest(modelo)

# 4. Linearity (RESET test)
resettest(modelo, power = 2:3, type = "fitted")
```

## 4. Regression with Categorical Variables

We can also use linear models to test differences across categories (which is mathematically equivalent to an independent t-test).

```{r categorical-regression}
# Fit a model predicting MPG based on transmission type (am)
modelo3 <- lm(mpg ~ am, data = mtcars) 
summary(modelo3)

# Perform a t-test to formally compare MPG across the two transmission types
t.test(mpg ~ am, data = mtcars)
```

## 5. Regression and Uncertainty

Let's visualize the fit of our model and understand the variance breakdown (SSR, ESS, TSS).

```{r uncertainty}
par(mfrow = c(1, 1))
plot(mtcars$wt, mtcars$mpg, main = "Actual vs Fitted Values",
     xlab = "Weight (1000 lbs)", ylab = "Miles Per Gallon (MPG)", pch = 19, col = "steelblue")
points(mtcars$wt, fitted(modelo), col = "firebrick", pch = 19)
legend("topright", legend = c("Actual", "Fitted"), col = c("steelblue", "firebrick"), pch = 19)

# Calculate Sum of Squared Residuals (SSR)
SRC <- sum((fitted(modelo) - mtcars$mpg)^2)

# Calculate Explained Sum of Squares (ESS)
SEC <- sum((fitted(modelo) - mean(mtcars$mpg))^2)

# Calculate Total Sum of Squares (TSS)
STC <- SRC + SEC

# Calculate R-squared manually
R2 <- SEC / STC
cat("Manually calculated R-squared:", R2, "\n")
```

### 5.1 Simulating Sampling Uncertainty

Let's consider 4 random samples of 10 cars to see how the regression line and confidence bands fluctuate around the true population line.

```{r sampling-simulation, fig.height=8, fig.width=8}
set.seed(123)
par(mfrow = c(2, 2)) # Set up a 2x2 grid for the 4 plots
for(i in 1:4){
  # Plot the full population
  plot(mpg ~ wt, data = mtcars,
       xlab = "Weight", ylab = "Miles Per Gallon",
       main = paste("Random Sample", i),
       ylim = c(min(mtcars$mpg), max(mtcars$mpg) + 0.3))
  
  # Add the true population regression line
  abline(modelo)
  
  if(i == 1){
    legend("topright", 
           pch = c(NA, NA, NA, 1, 16), 
           lty = c(1, 1, 2, NA, NA),
           col = c(1, 2, 2, 1, 2),
           legend = c("Population truth", "Sample estimate",
                      "Sample conf. band", "Unselected", "Sampled"),
           cex = 0.7, bty = "n")
  }
  
  # Randomly select 10 cars
  selected.cars <- sample(1:nrow(mtcars), 10)
  
  # Highlight the sampled points in red
  points(mtcars[selected.cars, "wt"], 
         mtcars[selected.cars, "mpg"], pch = 16, col = 2)
  
  # Fit a regression line using only the sample
  model.sel <- lm(mpg ~ wt, data = mtcars[selected.cars, ])
  abline(model.sel, col = 2)
  
  # Make a confidence band
  ww <- qt(0.975, 10 - 2)
  plot.x <- data.frame(wt = seq(1.513, 5.424, 0.1))
  plot.fit <- predict(model.sel, plot.x, level = 0.95, interval = "confidence", se.fit = TRUE)
  
  lines(plot.x$wt, plot.fit$fit[, 1] + ww * plot.fit$se.fit, col = 2, lty = 2)
  lines(plot.x$wt, plot.fit$fit[, 1] - ww * plot.fit$se.fit, col = 2, lty = 2)
}
```

## 6. Multiple Regression

We will generate a synthetic dataset for public policies to model `economic_output`.

```{r multiple-regression}
# Generate a synthetic dataset for public policies
set.seed(123)
n <- 1000
public_policies_data <- data.frame(
  region = paste("Region", 1:n),
  education_expenditure = runif(n, 5000, 15000),   # Education spending
  healthcare_expenditure = runif(n, 2000, 10000),  # Healthcare spending
  poverty_rate = runif(n, 5, 30)                   # Poverty rate (%)
)

# Calculate economic output based on true coefficients and random noise
public_policies_data$economic_output <- 10000 + 
  0.5 * public_policies_data$education_expenditure + 
  0.3 * public_policies_data$healthcare_expenditure - 
  200 * public_policies_data$poverty_rate + 
  rnorm(n, 0, 5000) 

# Fit a multiple regression model
modelo_mult <- lm(economic_output ~ education_expenditure + healthcare_expenditure + poverty_rate, 
                  data = public_policies_data)

# Summarize the multiple regression model
summary(modelo_mult)
```

### 6.1 Diagnostics for Multiple Regression

With multiple predictors, we also need to check for **multicollinearity**.

```{r multiple-diagnostics}
# 1. Absence of Multicollinearity (VIF)
# VIF values > 5 or 10 indicate problematic multicollinearity.
vif(modelo_mult)

# 2. Normality of Residuals
shapiro.test(residuals(modelo_mult))

# 3. Homoscedasticity 
bptest(modelo_mult)

# 4. Independence of Residuals
dwtest(modelo_mult)

# 5. Linearity
resettest(modelo_mult, power = 2:3, type = "fitted")
```

::: {.callout-tip title="Cross-References"}
* Return to [Topics overview](../../README.md)
:::
