---
title: "Data Lab: Ordered Logistic Regression"
format:
  html:
    code-fold: false
---

::: {.callout-info title="Lab Overview"}
In this practical lab, we will analyze data using an **Ordinal Logistic Regression** (Proportional Odds Model). We will fit the model, interpret the odds ratios, verify the proportional odds assumption, and visualize predicted probabilities.
:::

## 1. Data Loading and Preparation

We'll load the UCLA ordinal dataset, which studies students' intent to apply to graduate school.

```{r setup, message=FALSE, warning=FALSE}
# Load required libraries
library(MASS)     # Required for polr() function
library(ggplot2)
library(dplyr)
library(tidyr)
library(tibble)
library(reshape2)
library(foreign)

# Load the dataset
dat <- read.dta("https://stats.idre.ucla.edu/stat/data/ologit.dta")
```

::: {.callout-warning title="Fixing Factor Contrasts"}
When importing Stata `.dta` files, categorical variables are sometimes parsed as **ordered factors**. When R runs regressions on ordered factors, it automatically uses orthogonal polynomial contrasts, leading to `.L` (linear) and `.Q` (quadratic) suffixes in the output parameters which are hard to interpret. 

To fix this, we explicitly convert our predictors to standard, unordered factors.
:::

```{r data-prep}
# Convert predictors to standard unordered factors to prevent .L and .Q contrasts
dat$pared <- factor(dat$pared, ordered = FALSE)
dat$public <- factor(dat$public, ordered = FALSE)

# Inspect the structure of the dataset
summary(dat)

# Verify the ordinal nature of the dependent variable
str(dat$apply)

# Generate a cross-tabulation
xtabs(~apply + pared, data = dat)
xtabs(~apply + public, data = dat)
```

## 2. Model Fitting

We estimate the Proportional Odds Logistic Regression model using `polr`.

```{r model-fit}
# Estimate the model
m <- polr(apply ~ pared + public + gpa, data = dat, Hess = TRUE)

# View model summary
summary(m)
```

### 2.1 Significance Testing

`polr` does not provide p-values by default, so we calculate them using the t-values and the normal distribution.

```{r p-values}
# Calculate p-values
ctable <- coef(summary(m))
p <- pnorm(abs(ctable[, "t value"]), lower.tail = FALSE) * 2

ctable <- cbind(ctable, "p value" = p)
print(round(ctable, 3))
```

### 2.2 Interpretation with Odds Ratios

We transform log-odds to Odds Ratios (OR) and compute 95% confidence intervals.

```{r odds-ratios}
ci <- confint(m)
OR <- exp(coef(m))
ci_exp <- exp(ci)

# Prepare data for Forest Plot visualization
df_plot <- tibble(
  variable = rownames(ci_exp),
  OR = OR,
  lower = ci_exp[, 1],
  upper = ci_exp[, 2],
  color = c("red", "steelblue", "steelblue1") # Custom colors for variables
)
df_plot$variable <- factor(df_plot$variable, levels = rev(df_plot$variable))

print(df_plot)
```

## 3. Diagnostic: Proportional Odds Assumption

A key assumption of the ordinal model is that the relationship between each pair of outcome groups is the same (Parallel Slopes). We test this by comparing binary logits for different thresholds.

```{r proportional-odds}
dat$y_gt1 <- as.numeric(dat$apply != "unlikely")      # unlikely vs (somewhat + very)
dat$y_gt2 <- as.numeric(dat$apply == "very likely")   # (unlikely + somewhat) vs very

m1 <- glm(y_gt1 ~ pared + public + gpa, data = dat, family = binomial)
m2 <- glm(y_gt2 ~ pared + public + gpa, data = dat, family = binomial)

# Compare coefficients
df1 <- tibble(modelo = "y > 1", variable = names(coef(m1)), OR = exp(coef(m1)))
df2 <- tibble(modelo = "y > 2", variable = names(coef(m2)), OR = exp(coef(m2)))

bind_rows(df1, df2) %>%
  filter(variable != "(Intercept)") %>%
  ggplot(aes(x = OR, y = variable, color = modelo)) +
  geom_point(position = position_dodge(width = 0.5), size = 3) +
  labs(title = "Proportional Odds Assumption Check",
       subtitle = "Coefficients should be similar across models",
       x = "Odds Ratio", y = "Predictor") +
  theme_minimal()
```

## 4. Predictions and Visualizations

We simulate a grid of values to visualize the model's predicted probabilities across GPA ranges.

```{r predictions, fig.width=9, fig.height=6}
# Create a simulation grid
newdat <- data.frame(
  pared = factor(rep(0:1, 200), ordered = FALSE),
  public = factor(rep(0:1, each = 200), ordered = FALSE),
  gpa = rep(seq(from = 1.9, to = 4, length.out = 100), 4)
)

# Predict probabilities
newdat <- cbind(newdat, predict(m, newdat, type = "probs"))

# Reshape data to long format
lnewdat <- melt(newdat, id.vars = c("pared", "public", "gpa"),
                variable.name = "Level", value.name = "Probability")

lnewdat$Level <- factor(lnewdat$Level,
                        levels = c("unlikely", "somewhat likely", "very likely"),
                        labels = c("Unlikely", "Somewhat Likely", "Very Likely"))

# Plot the curves
ggplot(lnewdat, aes(x = gpa, y = Probability, colour = Level)) +
  geom_line(size = 1.2) +
  facet_grid(pared ~ public, labeller = label_both) + 
  scale_color_manual(values = c("Unlikely" = "red", 
                                "Somewhat Likely" = "steelblue", 
                                "Very Likely" = "steelblue1")) +
  labs(
    x = "Grade Point Average (GPA)",
    y = "Predicted Probability",
    colour = "Intent Level",
    title = "Predicted Probabilities of Applying to Graduate School",
    subtitle = "By GPA, Parental Education, and Institution Type"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    strip.text = element_text(face = "bold", size = 12),
    legend.position = "bottom",
    legend.title = element_text(face = "bold"),
    plot.title = element_text(face = "bold", size = 14, hjust = 0.5)
  )
```

## 5. Forest Plot

```{r forest-plot, fig.width=7, fig.height=4}
ggplot(df_plot, aes(x = OR, y = variable)) +
  geom_point(aes(color = color), size = 3) +
  geom_errorbarh(aes(xmin = lower, xmax = upper, color = color), height = 0.2) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "gray50") +
  scale_x_log10() +
  scale_color_identity() +
  labs(
    x = "Odds Ratio (OR) (Log Scale)",
    y = "Predictor Variables",
    title = "Forest Plot: Ordinal Logistic Regression"
  ) +
  theme_minimal(base_size = 13)
```

::: {.callout-tip title="Cross-References"}
* Return to [Topics overview](../../README.md)
:::
