---
title: "Data Lab: Logistic Regression (Binomial Model)"
format:
  html:
    code-fold: false
---

::: {.callout-info title="Lab Overview"}
In this practical lab, we will analyze data using a **Binomial Logistic Regression Model**. We will explore the data, fit the model, verify assumptions (such as linearity of the logit and multicollinearity), and visualize predicted probabilities across different profiles.
:::

## 1. Data Loading and Preparation

We load the UCLA binary dataset which studies how GRE scores (renamed to `puntaje_ICFES`), GPA, and undergraduate institution rank influence admission to postgraduate programs. The response variable, `admit`, is binary.

```{r setup, message=FALSE, warning=FALSE}
# Load required libraries
library(aod)
library(ggplot2)
library(dplyr)
library(tidyr)
library(car) # Required for logistic regression assumption tests (VIF, Box-Tidwell)

# Load the dataset
dt <- read.csv("https://stats.idre.ucla.edu/stat/data/binary.csv")

# Rename 'gre' to 'puntaje_ICFES' for local context
names(dt)[names(dt) == "gre"] <- "puntaje_ICFES"

# Convert rank into an explicitly ordered factor
dt$rank_cat <- NA
dt$rank_cat[dt$rank == 1] <- "rango_1"
dt$rank_cat[dt$rank == 2] <- "rango_2"
dt$rank_cat[dt$rank == 3] <- "rango_3"
dt$rank_cat[dt$rank == 4] <- "rango_4"

dt$rank_ord <- factor(dt$rank_cat, 
                      levels = c("rango_1", "rango_2", "rango_3", "rango_4"), 
                      ordered = TRUE)

# Format the binary response variable 'admit' into a descriptive factor
dt <- dt %>%
  mutate(admit = factor(admit, levels = c(0, 1), labels = c("No admitido", "Admitido")))
```

## 2. Exploratory Data Analysis (EDA)

Before running the model, let's explore the relationships in our data.

```{r eda}
summary(dt)

# Cross-tabulation between admission and university rank
xtabs(~admit + rank_ord, data = dt)

# Comparative Histogram for ICFES scores
hist(dt$puntaje_ICFES, 
     col = rgb(0, 0, 1, 0.5), 
     main = "Distribución Comparativa de Puntajes ICFES", 
     xlab = "Puntaje", 
     ylab = "Frecuencia",
     xlim = c(200, 800)) 
```

```{r boxplots, fig.width=8, fig.height=4}
p1 <- ggplot(dt, aes(x = admit, y = puntaje_ICFES, fill = admit)) +
  geom_boxplot() +
  scale_fill_manual(values = c("steelblue", "red")) +
  labs(title = "ICFES por admisión", x = "Admisión", y = "ICFES") +
  theme_minimal() +
  theme(legend.position = "none")

p2 <- ggplot(dt, aes(x = admit, y = gpa, fill = admit)) +
  geom_boxplot() +
  scale_fill_manual(values = c("steelblue", "red")) +
  labs(title = "GPA por admisión", x = "Admisión", y = "GPA") +
  theme_minimal() +
  theme(legend.position = "none")

library(patchwork) # To display side by side
p1 + p2
```

## 3. Model Fitting

We estimate the binomial logit model using the Generalized Linear Model (`glm`) function with a logistic link.

```{r model-fit}
# Estimate the binomial logit model
logit <- glm(admit ~ puntaje_ICFES + gpa + rank_ord, data = dt, family = "binomial")

# View the summary
summary(logit)
```

> **Interpretation:** The estimated coefficients are in log-odds units. For example, for each additional unit in the ICFES score, the log-odds of admission increase by about 0.002. For each additional unit in GPA, the log-odds increase by 0.804. Since `rank_ord` is an ordered factor, R outputs polynomial contrasts (`.L`, `.Q`, `.C`) to capture linear and non-linear trends across ranks.

### 3.1 Odds Ratios

To interpret the coefficients properly, we transform them from log-odds into Odds Ratios (OR).

```{r odds-ratios}
exp(cbind(OR = coef(logit), confint(logit)))
```

## 4. Verification of Logistic Regression Assumptions

Logistic regression doesn't assume normality of residuals or homoscedasticity like OLS, but it has its own strict assumptions.

```{r diagnostics}
# 1. Absence of Multicollinearity (VIF)
# Values near 1 are excellent.
vif(logit)

# 2. Absence of Strongly Influential Outliers
# Points significantly higher than the rest warrant further investigation.
plot(logit, which = 4, id.n = 3)
```

## 5. Predictions and Visualizations

We calculate predicted probabilities of admission holding GPA at its mean, across all 4 institution ranks and varying ICFES scores.

```{r predictions, fig.width=8, fig.height=5}
# Create simulation grid
n_dt2 <- with(dt, data.frame(
  puntaje_ICFES = rep(seq(from = 200, to = 800, length.out = 100), 4), 
  gpa = mean(gpa, na.rm = TRUE), 
  rank_ord = factor(rep(c("rango_1", "rango_2", "rango_3", "rango_4"), each = 100), 
                    levels = c("rango_1", "rango_2", "rango_3", "rango_4"), 
                    ordered = TRUE)
))

# Predict values on link scale (log-odds) and compute standard errors
n_dt3 <- cbind(n_dt2, predict(logit, newdata = n_dt2, type = "link", se = TRUE))

# Transform into probabilities using plogis
n_dt3 <- within(n_dt3, {
  PredictedProb <- plogis(fit)
  LL <- plogis(fit - (1.96 * se.fit))
  UL <- plogis(fit + (1.96 * se.fit))
})

# Plot the predicted probability curves
ggplot(n_dt3, aes(x = puntaje_ICFES, y = PredictedProb)) +
  geom_ribbon(aes(ymin = LL, ymax = UL, fill = rank_ord), alpha = 0.2) +
  geom_line(aes(colour = rank_ord), size = 1) +
  scale_fill_manual(values = c("steelblue4", "steelblue2", "indianred2", "indianred4")) +
  scale_colour_manual(values = c("steelblue4", "steelblue2", "indianred2", "indianred4")) +
  labs(x = "Puntaje ICFES (GRE)", y = "Probabilidad Predicha de Admisión", 
       fill = "Rank", colour = "Rank",
       title = "Probabilidad de Admisión por Puntaje y Rango de Universidad") +
  theme_bw(base_size = 12) +
  theme(legend.position = "bottom",
        legend.title = element_text(face = "bold"))
```

### 5.1 Specific Population Profiles

We can evaluate specific cases of admission probability by constructing concrete student profiles.

```{r profiles}
# Create concrete population profiles
perfil_A <- data.frame(puntaje_ICFES = 800, gpa = 4.0, 
                       rank_ord = factor("rango_4", levels = c("rango_1", "rango_2", "rango_3", "rango_4"), ordered = TRUE))

perfil_B <- data.frame(puntaje_ICFES = 600, gpa = 3.0, 
                       rank_ord = factor("rango_1", levels = c("rango_1", "rango_2", "rango_3", "rango_4"), ordered = TRUE))

# Predict linear index (log-odds)
pred_A <- predict(logit, newdata = perfil_A, type = "link", se.fit = TRUE)
pred_B <- predict(logit, newdata = perfil_B, type = "link", se.fit = TRUE)

# Convert log-odds into bounded probabilities
cat("Probabilidad Perfil A (Top ICFES/GPA, Rango 4):", round(plogis(pred_A$fit), 3), "\n")
cat("Probabilidad Perfil B (Mid ICFES/GPA, Rango 1):", round(plogis(pred_B$fit), 3), "\n")
```

::: {.callout-tip title="Cross-References"}
* Return to [Topics overview](../../index.qmd)
:::
